Estimate of the Neural Net Dimension Using Algebraic Topology and Lie Theory\thanks{The code can be found at: \href{https://github.com/karhunenloeve/Irhe}{https://github.com/karhunenloeve/Irhe}.}

### 2020-06-17 llncs2e/paper12.tex
paper12$

\subsubsection*{Differential geometry in neural nets} The manifold of a dense neural net is in most cases Euclidean. Recently there have been results in the direction of other manifolds on which neural nets can operate. The spherical neural nets operate on a manifold with a commutative group structure \cite{CohenGKW18,EstevesAMD20}.

@@

\subsubsection{Persistence landscapes} Persistence landscapes give a statistical summary of the topology of a set of points embedded in a given topological manifold \cite{bubenik2015statistical,bubenikD17}. Looking at the points $(\epsilon_i,\epsilon_{j+1}) \in \overline{\mathbb{R}^2}$ on the $k$-th persistence diagram $\text{PH}_{k}(X)$, one associates a piecewise linear function $\lambda^{\epsilon_i}_{\epsilon_{j+1}}: \mathbb{R} \rightarrow [0,\infty)$ with those points:
\begin{align}
    &\text{If } x \not\in (\epsilon_i,\epsilon_{j+1}), \quad \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = 0,\\
    &\text{if } x \in \left(\epsilon_{i},(\epsilon_i + \epsilon_{j+1})/2 \right], \quad \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = x - \epsilon_{i} \; \text{and}\\
    &\text{if } x \in \left((\epsilon_i + \epsilon_{j+1})/2,\epsilon_{j+1} \right), \quad \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = \epsilon_{j+1} - x.
\end{align}
The summaries for the general persistence diagram are the disjoint union of the $k$-th persistence diagrams $\text{PH}(X) := \coprod_{i=0}^{k} \text{PH}_i(X)$. A persistence landscape $\text{PL}(X)$, contains the birth-death pairs $(\epsilon_i,\epsilon_{j+1})$, for an $i,j \in \{1, \cdots, n\}$ and is the function sequence $\Lambda_k : \mathbb{R} \rightarrow [0,\infty)$ for a $k \in \mathbb{N}$, where $\Lambda_k(x)$ denotes the $k$-th greatest value of $\lambda^{\epsilon_i}_{\epsilon_{j+1}}(x)$, see Fig. \ref{figure1} c)4. Thus, $\Lambda_k(x) = 0$ if $k > n$. $\text{PL}(X)$ lies in a completely normed vector space, suited for statistical computations.

@@

\section{Counting Betti numbers}
Based on our assumption that data lies on a connected Abelian Lie group, we can decompose the data space into a product of realaxes and tori. In the following we refer to this Lie group as $G \simeq \mathbb{R}^p \times \mathbb{T}^q$. By this decomposition there is also an isomorphism of the homology groups of $G$ with those of $\mathbb{R}^p \times \mathbb{T}^q$, so that $H_k(G) \simeq H_k(\mathbb{R}^p \times \mathcal{S}^{1}_{1}\times\cdots\times\mathcal{S}^{1}_{q})$ with $p+q = \dim G$. We demand from a neural net that these homology groups can all be approximated, i.e. that all topological structure of the input space can be represented sufficiently well.

One might wonder to what extent the homology groups of a simplicial complex can be used to estimate the homology groups of the manifolds underlying the data. In this section we denote the simplicial homology groups as $H^{\Delta}_{\bullet}(X)$. The $\bullet$ is a proxy for all possible indices. Through the construction of continuous mappings, so-called singular simplices, from a simplex into a topological space, the singular homology groups $H_{\bullet}(X)$ can be constructed. We refer to Hatcher \cite[p.~102]{hatcher2002algebraic} for a detailed introduction with proofs. This allows to assign Abelian groups to any topological space. If one constructs smooth mappings instead of continuous ones, one gets smooth singular homology groups $H^{\infty}_{\bullet}(X)$, see for a proof \cite[pp.~473]{lee2013smooth}. Finally a chain complex can be formed on a smooth manifold over the $p$-forms on $X$ and from this the so-called De Rham cohomology can be defined $H_{\text{dR}}^{\bullet}(X)$, for a proof we refer to \cite[pp.~440]{lee2013smooth}. In summary, we have homology theories for simplicial complexes, topological spaces, a smooth variant on topological spaces and on smooth manifolds. We consider simplicial homology to be determined on the basis of a simplicial complex, which is merely a rough approximation of the triangulation of the underlying topological space. However, we can draw conclusions about a possible smooth manifold, since the discussed homology groups all have isomorphisms to each other known as the De Rham Theorem, see \cite[pp.~106, pp.~467]{hatcher2002algebraic,lee2013smooth}. This legitimizes our approach.

@@

\section{Experimental setting}
We train an auto-encoder using as input heavily noisy images and map them to their noiseless original. We perturb each input vector $\textbf{x}^{(0)}$ with Gaussian noise $\epsilon \sim \mathcal{N}(0,\sigma^2 \mathbf{I})$, such that the input is weighted $0.5 \cdot \textbf{x}^{(0)} + 0.5 \cdot \epsilon$. The embedding dimension is increased for each experiment by $2$, i.e. $2,4,6, \cdots, 784$. $392$ neural nets have been trained $\times 100$. Each line of Fig. \ref{figure2} c), d) represents one neural architecture as averaged loss function.
\input{figure2.tex}
\subsubsection*{Persistence landscapes} We use the Delaunay-complex for the filtration, according to Melodia et al. \cite{lume}. The maximum $\alpha$-square is set to $e^{12}$. The maximal expansion dimension of the simplicial complex is set to $10$. The maximal edge length is set to $0.1$. The persistent landscapes are smoothened by a Gaussian filter $G(x) = 1 / \sqrt{2 \pi \sigma^2} \cdot e^{- x^2 / 2 \sigma^2}$ with $\sigma = 2$ for visualization purposes. We implement the persistence diagrams and persistence landscapes using the \texttt{GUDHI} library \cite{gudhi:urm}. Persistent homology is computed on a \texttt{NVIDIA Quadro P4000}.
\subsubsection*{Hyperparameters} The batch size is set to $128$, which should help to better recognize good parametrization. Higher batch size causes explosions of the gradient, which we visualize in Fig. \ref{figure2} c), d). For optimization we use \texttt{ADAM} \cite{KingmaB14} and a learning rate of $10^{-4}$. The dataset is randomly shuffled. We implement in \texttt{Python} $3.7$. We use \texttt{Tensorflow} as backend \cite{tensorflow2015-whitepaper} with \texttt{Keras} as wrapper \cite{chollet2015keras}. The neural training is conducted on \texttt{Intel Core i}$7$ $9700$\texttt{K} processors.
\input{figure3.tex}
\subsubsection*{Neural net} We use $\text{L}(\textbf{x}^{(0)}, \textbf{y}) = 1/n \sum_{i=1}^{n} (\textbf{x}^{(0)}_i-\textbf{y}_i)^2$ as loss function. The sigmoid activation function $\sigma(\mathbf{x}^{(l)}) = 1 / 1 + e^{-\mathbf{x}^{(l)}}$ is applied throughout the net to yield the structure of a smooth manifold. Dense neural nets are used with bias term. The datasets are flattened from $(32,32,3)$ into $(3072)$. The first layer is applied to the flattened dataset, which is also adapted by back propagation, i.e. $\varphi^{(0)}(x^{(0)}) =$ $\sigma(\sum_{i=1}^{n} \textbf{W}_{ji}^{(1)} \textbf{x}_{i}^{(0)} + \textbf{b}_{i}^{(1)})$. It transfers the data to the desired embedding dimension. The subsequent layers are implemented as residual invertible layers following Dinh et al. \cite{DinhSB17}. We use invertible architectures to stay on the same differentiable structure during learning. A total of $5$ hidden layers are used, all with the same embedding dimension.

\subsubsection*{Results} In Fig. \ref{figure2} c) d) we colored the neural nets according to the chosen embedding, such that the lines show the following dimensions ${\color{red1}\bullet} = [2, \cdots, 148]$, ${\color{red2}\bullet} = [150, \cdots, 198]$, ${\color{red3}\bullet} = [200, \cdots, 292]$ and ${\color{red4}\bullet} = [\dim U \approx 294, \cdots, 784]$.  The embedding dimension can be taken from Tab. \ref{figure3}, due to the trivial invertibility of our neural nets this integer must be doubled, see \cite{DinhSB17}. All models below our dimensional treshold $\dim U$ show drastic explosions of the gradient, see Fig. \ref{figure2} c) d). The models above the treshold remain stable. A single net still shows an explosive gradient colored ${\color{red4}\bullet}$ in Fig. \ref{figure2} d). The net showing the deflection of gradient is very close to our threshold, identified as $292$ dimensional embedding. The closer we come to the treshold, the sparser the fluctuations of loss. We justified that the minimal embedding dimension does not lie below our treshold.

\section{Conclusion}
Based on the theory of Lie groups and persistent homology a method for neural nets has been developed to parameterize them using the assumption that the data lies on or near by some connected commutative Lie group. This forms an approximate solution for a special case of learning problems, allowing to restrict the domain to a commutative group beeing a connected smooth manifold.

Applying Künneth's theorem, the homology groups of the topological factor spaces could be connected with the ones of their product space. Using persistence landscapes the elements originating from some homology groups on the filtration were estimated. With numerical experiments we predicted near ideal embedding dimensions and could confirm, according to Tab. \ref{figure3}, that such a neural embedding delivers a loss function on the validation data set with the smallest fluctuations and the best reconstruction results. We pose following open research questions:
\begin{itemize}
    \item The neural layers do not have the explicit structure of an Abelian Lie group. How can spherical CNNs \cite{CohenGKW18} be used to always represent a product space $\mathbb{R}^p \times \mathbb{T}^q$ from layer to layer in order to operate on the proposed Lie group?
    \item This approach can be applied to any kind of manifold, as far as its minimal representation is known. Using decompositions, one may generalize this result. What decomposition allows to neglect the connectedness?
\end{itemize}

>>>>

\subsubsection*{Differential geometry in neural nets} The manifold of a dense neural net is in most cases Euclidean. Recently there have been results in the direction of other manifolds on which neural nets can operate. The spherical neural nets operate on a manifold with a commutative group structure \cite{CohenGKW18}.

@@

\subsubsection{Persistence landscapes} Persistence landscapes give a statistical summary of the topology of a set of points embedded in a given topological manifold \cite{bubenik2015statistical}. Looking at the points $(\epsilon_i,\epsilon_{j+1}) \in \overline{\mathbb{R}^2}$ on the $k$-th persistence diagram $\text{PH}_{k}(X)$, one associates a piecewise linear function $\lambda^{\epsilon_i}_{\epsilon_{j+1}}: \mathbb{R} \rightarrow [0,\infty)$ with those points:
\begin{align}
    &\text{If } x \not\in (\epsilon_i,\epsilon_{j+1}), \quad \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = 0,\\
    &\text{if } x \in \left(\epsilon_{i},(\epsilon_i + \epsilon_{j+1})/2 \right], \quad \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = x - \epsilon_{i} \; \text{and}\\
    &\text{if } x \in \left((\epsilon_i + \epsilon_{j+1})/2,\epsilon_{j+1} \right), \quad \lambda^{\epsilon_i}_{\epsilon_{j+1}}(x) = \epsilon_{j+1} - x.
\end{align}
The summaries for the general persistence diagram are the disjoint union of the $k$-th persistence diagrams $\text{PH}(X) := \coprod_{i=0}^{k} \text{PH}_i(X)$. A persistence landscape $\text{PL}(X)$, contains the birth-death pairs $(\epsilon_i,\epsilon_{j+1})$, for an $i,j \in \{1, \cdots, n\}$ and is the function sequence $\Lambda_k : \mathbb{R} \rightarrow [0,\infty)$ for a $k \in \mathbb{N}$, where $\Lambda_k(x)$ denotes the $k$-th greatest value of $\lambda^{\epsilon_i}_{\epsilon_{j+1}}(x)$, see Fig. \ref{figure1} c)4. Thus, $\Lambda_k(x) = 0$ if $k > n$. $\text{PL}(X)$ lies in a completely normed vector space, suited for statistical computations.

@@

\section{Counting Betti numbers}
According to our main assumption, we seek a decomposition of the data manifold into a product of realaxes and tori. We refer to this Lie group as $G \simeq \mathbb{R}^p \times \mathbb{T}^q$. Thus, there is also an isomorphism of the homology groups of $G$ with those of $\mathbb{R}^p \times \mathbb{T}^q$, so that $H_k(G) \simeq H_k(\mathbb{R}^p \times \mathcal{S}^{1}_{1}\times\cdots\times\mathcal{S}^{1}_{q})$ with $p+q = \dim G$. We demand from a neural net that these homology groups can all be approximated, i.e. that all topological structure of the input space can be represented sufficiently well.

@@

One might wonder to what extent homology groups of a simplicial complex can be used to estimate homology groups of smooth manifolds. Through construction of continuous maps from a simplex into a topological space, so-called singular simplices, the singular homology groups $H_{k}(X)$ are obtained. We refer to Hatcher \cite[p.~102]{hatcher2002algebraic} for a proof. This allows to assign Abelian groups to any topological space. If one constructs smooth mappings instead of continuous ones, one gets smooth singular homology groups $H^{\infty}_{k}(X)$, consider \cite[pp.~473]{lee2013smooth} for a proof. Finally a chain complex can be formed on a smooth manifold over the $p$-forms on $X$ and from this the so-called De Rham cohomology can be defined $H_{\text{dR}}^{k}(X)$, for a proof we refer to \cite[pp.~440]{lee2013smooth}. In summary, we have homology theories for simplicial complexes, topological spaces and on smooth manifolds. Simplicial homology is determined by a simplicial complex, which is merely a rough approximation of the triangulation of the underlying topological space. However, we can draw conclusions about a possible smooth manifold, since the discussed homology groups all have isomorphisms to each other known as the De Rham Theorem, see \cite[pp.~106, pp.~467]{hatcher2002algebraic,lee2013smooth}. This legitimizes our approach.

\section{Experimental setting}
\input{figure2.tex}
We train an auto-encoder using as input heavily noisy images and map them to their noiseless original. We perturb each input vector $\textbf{x}^{(0)}$ with Gaussian noise $\epsilon \sim \mathcal{N}(0,\sigma^2 \mathbf{I})$, such that the input is weighted $0.5 \cdot \textbf{x}^{(0)} + 0.5 \cdot \epsilon$. The layer size is increased for each experiment by $2$, i.e. $2,4,6, \cdots, 784$. $392$ neural nets have been trained $\times 100$. Each line of Fig. \ref{figure2} c), d) represents one neural architecture as averaged loss function.\\[0.4cm]
\textbf{Persistence Landscapes:} We use the Delaunay-complex for the filtration, according to Melodia et al. \cite{lume}. The maximum $\alpha$-square is set to $e^{12}$. The maximal expansion dimension of the simplicial complex is set to $10$. The maximal edge length is set to $0.1$. The persistent landscapes are smoothened by a Gaussian filter $G(x) = 1 / \sqrt{2 \pi \sigma^2} \cdot e^{- x^2 / 2 \sigma^2}$ with $\sigma = 2$ for visualization purposes. We implement the persistence diagrams and persistence landscapes using the \texttt{GUDHI} library \cite{gudhi:urm}. Persistent homology is computed on a \texttt{NVIDIA Quadro P4000}.\\[0.4cm]
\textbf{Neural nets:} We use $\text{L}(\textbf{x}^{(0)}, \textbf{y}) = 1/n \sum_{i=1}^{n} (\textbf{x}^{(0)}_i-\textbf{y}_i)^2$ as loss function. The sigmoid activation function $\sigma(\mathbf{x}^{(l)}) = 1 / 1 + e^{-\mathbf{x}^{(l)}}$ is applied throughout the net to yield the structure of a smooth manifold. Dense neural nets are used with bias term. The datasets are flattened from $(32,32,3)$ into $(3072)$. The first layer is applied to the flattened dataset, which is also adapted by back propagation, i.e. $\varphi^{(0)}(x^{(0)}) =$ $\sigma(\sum_{i=1}^{n} \textbf{W}_{ji}^{(1)} \textbf{x}_{i}^{(0)} + \textbf{b}_{i}^{(1)})$. It transfers the data to the desired embedding dimension. The subsequent layers are implemented as residual invertible layers following Dinh et al. \cite{DinhSB17}. We use invertible architectures to stay on the same differentiable structure during learning. A total of $5$ hidden layers are used, all with the same embedding dimension. The batch size is set to $128$, which should help to better recognize good parametrization. Higher batch size causes explosions of the gradient, which we visualize in Fig. \ref{figure2} c), d). For optimization we use \texttt{ADAM} \cite{KingmaB14} and a learning rate of $10^{-4}$. The dataset is randomly shuffled. We implement in \texttt{Python} $3.7$. We use \texttt{Tensorflow} as backend \cite{tensorflow2015-whitepaper} with \texttt{Keras} as wrapper \cite{chollet2015keras}. The training is conducted on \texttt{Intel Core i}$7$ $9700$\texttt{K} processors.\\[0.4cm]
\textbf{Results:} In Fig. \ref{figure2} c) d) we colored the neural nets according to the chosen embedding, such that the lines show the following dimensions ${\color{red1}\bullet} = [2, \cdots, 148]$, ${\color{red2}\bullet} = [150, \cdots, 198]$, ${\color{red3}\bullet} = [200, \cdots, 292]$ and ${\color{red4}\bullet} = [\dim U \approx 294, \cdots, 784]$.  The embedding dimension can be taken from Tab. \ref{figure3}.\footnote{Due to the construction of invertibility the treshold must be doubled, see \cite{DinhSB17}.} All models below our dimensional treshold $\dim U$ show drastic explosions of the gradient, see Fig. \ref{figure2} c) d). The models above the treshold remain stable. A single net still shows an explosive gradient colored ${\color{red4}\bullet}$ in Fig. \ref{figure2} d). The net showing the deflection of gradient is very close to our threshold, identified as $292$ dimensional embedding. The closer we come to the treshold, the sparser the fluctuations of loss.
\input{figure3.tex}

\section{Conclusion}
Based on the theory of Lie groups and persistent homology a method for neural nets has been developed to parameterize them using the assumption that the data lies on or near by some connected commutative Lie group. This forms an approximate solution for a special case of learning problems.

Applying Künneth's theorem, the homology groups of the topological factor spaces could be connected with the ones of their product space. Using persistence landscapes the elements originating from some homology groups on the filtration were estimated. With numerical experiments we predicted near ideal embedding dimensions and could confirm, see Fig. \ref{figure2}, that a neural embedding above the treshold delivers a loss function on the validation data set with small fluctuations and best reconstruction results. We pose following open research questions:
\begin{itemize}
    \item The neural layers do not have the explicit structure of an Abelian Lie group. How can spherical CNNs \cite{CohenGKW18} be used to always represent a product space $\mathbb{R}^p \times \mathbb{T}^q$ from layer to layer in order to operate on the proposed Lie group?
    \item This approach can be applied to any kind of manifold, as far as its minimal representation is known. Using decompositions, one may generalize this result. What decomposition allows to neglect the connectedness?
\end{itemize}
